CrypTool Script \\ Mathematics and Cryptography

### 2003-03-10 script/en/chapters/primes.tex
no message

\newpage
\section{Prime numbers}
\hypertarget{Kapitel_2}{}
(Bernhard Esslinger, besslinger@web.de, May 1999, Updates Nov. 2000, December 2001, February 2003)

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid
\footnote{Euclid,
a Greek mathematician of 4th and 3rd century before Christ. He worked at the
Egyptian academy of Alexandria and wrote ``The Elements'', the most well 
known systematically textbook of the Greek mathematics.} 
knew this in the third century before Christ.
\begin{theorem}[Euclid]\label{thm-pz-euklid}\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid’s Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as followed
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than any previously existing amount
of prime numbers.}
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

% --------------------------------------------------------------------------
\hypertarget{MersenneNumbers01}{}
\subsubsection{Special number types -- Mersenne numbers} \index{ Mersenne numbers}

@@

\index{Number!Mersenne}\index{Mersenne!number}
\index{Mersenne!theorem} Even Mersenne knew that not all Mersenne numbers
are prime (see exponent $p = 11$). A prime Mersenne number \index{Mersenne!prime number} is called Mersenne prime number.  
However, he is to be thanked for the interesting conclusion that a number of the form $2^n-1$ is not a prime number
if $n$ is a composite number:

@@

From \begin{eqnarray*} (x^r-1)((x^r)^{s-1} + (x^r)^{s-2} + \cdots + x^r +1) & =
&  ((x^r)^s + (x^r)^{s-1} + (x^r)^{s-2} + \cdots + x^r) \\ &  & -((x^r)^{s-1} +
(x^r)^{s-2} + \cdots + x^r +1)  \\ & = & (x^r)^s -1 = x^{rs } -1,
\end{eqnarray*} we conclude \[ 2^{n_1 n_2} - 1 = (2^{n_1} -1)((2^{n_1})^{n_2 -1}
+ (2^{n_1})^{n_2 -2} + \cdots + 2^{n_1} + 1). \]
Because $ 2^n - 1 $ is a prime number, one of the above two factors on the
right-hand side must be equal to 1. This is the case if and only if $ n_1 =1 $
or $ n_2 =1$. But this contradicts our assumption. Therefore the assumption is
false. This means that there exists no composite number $ n, $ such that $ 2^n -
1 $ is a prime.
\end{Proof} 
\vskip + 5pt
Unfortunately this theorem only applies in one direction (the inverse does not
apply, no equivalence): see the above example $2^{11}-1, $ where $11$ is prime.

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before \index{Lucas
  Edouard} Edouard Lucas (1842-1891) proved that this number is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson showed in 1903 which factors make up this composite number: $$ 2^{67} -1
=147, 573, 952, 588, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorisation 
\index{factorisation} (expression as a product of factors) 
\footnote{%
Using CrypTool\index{CrypTool} v1.3 you can factorize numbers in the 
following way: menu {\bf Indiv. Procedures \textbackslash RSA Demonstration 
\textbackslash Factorisation of a Number}.
}
of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers \index{Mersenne!prime number} 
have the exponents 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377; ~
6,972,593; \\13,466,917.
\end{array}
$$
Thus $39$ Mersenne prime numbers are currently known. For the first 38 Mersenne prime numbers we know that this list is complete. The exponents until the 39th Mersenne prime numbers are not yet checked completely.

@@

These numbers can be found at the following URLs:

{\href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html}}
\vskip - 4pt \noindent
\begin{quote}
(The supercomputer manufacturer SGI Cray Research not only employed brilliant
mathematicians but also used the prime number tests as benchmarks for their
machines.)
\end{quote}
{\href{http://www.utm.edu/}{\tt http://www.utm.edu/}}
\vskip - 4pt \noindent
\begin{quote}
(At the university of Tennessee you will find extensive research results about
prime numbers.)
\end{quote}

\vskip +15 pt
\paragraph{M-37 -- January 1998}\index{Mersenne!prime number!M-37}\mbox{}
% \\
The 37th Mersenne prime,
$$ 2^{3,021,377} - 1. $$
was found in January 1998 and has 909,526
digits in the decimal system, which corresponds to 33 pages in the newspaper!

\vskip +15 pt
\paragraph{M-38 -- June 1999}\index{Mersenne!prime number!M-38}\mbox{}
% \\
The 38th Mersenne prime, called M-38,
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

\vskip +15 pt
\paragraph{M13466917 -- December 2001} \index{Mersenne!prime number!M-39}\mbox{}
% \\
This number was discovered as 39th Mersenne prime (and already called M-39, despite it is not proven yet, whether there do not exist further Mersenne prime numbers between M-38 und M13466917), $$2^{13,466,917}-1,$$ at December 6, 2001 –- more exactly, the verification of this number, found at November 14, 2001 by the Canadian student Michael Cameron, was successfully finished. This number has about 4 million decimal digits (exactly 4,053,946 digits).
Trying only to print this number $(924947738006701322247758 \cdots 1130073855470256259071)$ in the Financial Times would need around 200 pages.

\vskip +15 pt
\paragraph{GIMPS}\index{GIMPS}\mbox{}
% \\
Discovering the 39th Mersenne prime the GIMPS project (Great Internet Mersenne Prime Search) founded in 1996 already discovered for the 5th time the greatest Mersenne number to be proven being prime
({\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}).

@@

\hypertarget{FermatNumbers01}{}
In the 17th century, Fermat wrote to Mersenne that he presumed that all numbers
of the form $$ F(n) = 2^{2^n} + 1 $$ are prime for all whole numbers $n$ greater
than or equal to $0$ (\hyperlink{FermatNumbers02}{see below}) 
\index{Zahlen!Fermatzahlen}\index{Fermat!Fermatzahl}

@@

Detailed descriptions of tests for finding out whether a number is prime can be
found on Web sites such as:

{\href{http://www.utm.edu/research/primes/mersenne.shtml}{\tt
http://www.utm.edu/research/primes/mersenne.shtml}} and

{\href{http://www.utm.edu/research/primes/prove/index.html}{\tt
http://www.utm.edu/research/primes/prove/index.html}}.

@@

Ideas for this can be found at \\
{\href{http://www.utm.edu/research/primes/notes/faq/p_n.html}
 {\tt http://www.utm.edu/research/primes/notes/faq/p\_n.html}}.

@@

\item Mersenne numbers  $f(n) = 2^n - 1$: \\
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems to deliver 
    relatively large prime numbers but - as for $n=11$ [$f(n)=2,047$] - it is 
    repeatedly the case that the result is not prime.
    Today, all the Mersenne primes having less than 2,000,000 digits are known\\
    ({\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}).

@@

    Within the project ``Distributed Search for Fermat Number Dividers'' offered by
    Leonid Durman there is also progress in finding new monster primes
    ({\href{http://www.fermatsearch.org/}
       {\tt http://www.fermatsearch.org/}}).
    This website has conjectured webpages in Russian, Italian and German.

@@

    The Fermat number
    $$ F[2,145,351] = 2^(2^{2,145,351}) + 1 $$ 
    is divisible by the prime
    $$ p = (3*2^{2,145,353} + 1) $$ \\
    This prime is the largest known prime generalized Mersenne number. 
    It was discovered only a few days after Michael Angel discovered the largest
    generalized Fermat number (\hyperlink{generalizedFermatprimes}{see below}).
    This moved GIMPS first prime discovery, M-35 = M1398269, into 7th place.

@@

    More details are in\\
    {\href{http://www.fermatsearch.org/history/cosgrave_record.htm/}
       {\tt http://www.fermatsearch.org/history/cosgrave\_record.htm}}.

%    ({\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
%           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}).
% M.E. ist die Zahl auf seiner Webseite zu groß, da M-39 "nur" 4 Mio. Stellen hat !
%    Heute kennt man alle Fermatschen Primzahlen mit bis zu 2.000.000.000 ??? 
%    Dezimalstellen. \\

\vskip +15 pt
\hypertarget{generalizedFermatprimes}{}
\item Generalized Fermat numbers    $F(b,m) = b^{2^m} + 1$: \\
\index{Fermat numbers}
    Generalized Fermat numbers are more numerous than Mersenne numbers at equal size 
    and many of them are waiting to be discovered to fill the big gaps between the 
    Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the representation is not 
    limited to the base 2, can be tested at almost the same speed than a 
    Mersenne number.

@@

    Using this program at February 16, 2003 Michael Angel discovered the largest of them 
    till then with 628,808 digits, which so became the 5th largest known prime number:
    $$ b^{2^{17}} + 1  =  62722^{131,072} + 1. $$ 
    This moved GIMPS first prime discovery, M-35 = M1398269, into 6th place.

   ({\href{http://www.prothsearch.net/index.html}
    {\tt http://www.prothsearch.net/index.html}})

   ({\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
    {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}})

@@

  \item \index{Euclidean number} {\em Euclidean numbers} $e_n = e_0 \cdot e_1
\cdots e_{n-1} + 1$  with $n$ greater than or equal to $1$ and $e_0 := 1$.
$e_{n-1}$ is not the $(n-1)$th prime number, but the number previously found
here.
    Unfortunately this formula is not open but recursive.
    The sequence starts with

@@

$e_9, \cdots, e_{17}$ are also composite, which means that this formula is not
particularly useful. Comment: However, what is special about all these numbers
is that any pair of them does not have a common factor other than $1$. They are
therefore \index{Prime number!relative} {\em relatively prime}.

@@

The first $40$ values are prime numbers (which have the obvious regularity that
their difference starts with $2$ and increases by $2$ each time), but the $41$th
and $42$th values are not prime numbers.     It is easy to see that $f(41)$ hl
cannot be a prime number:
    $f(41) = 41^2 + 41 + 41 = 41 (41 + 1 + 1) = 41 \cdot 43$.

@@

@@

% --------------------------------------------------------------------------
\subsection{Density and distribution of the primes}

@@

Under the heading {\em How Many Primes Are There} under

{\href{http://www.utm.edu/research/primes/howmany.shtml}{\tt
http://www.utm.edu/research/primes/howmany.shtml}}.

you will find numerous other details.

Using the following Web site:

{\href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{\tt
http://www.math.Princeton.EDU/\~{}arbooker/nthprime.html}}

you can easily determine $PI(x)$.

The \textbf{distribution} of primes displays several irregularities for which no
€system€ has yet been found. (On the one hand, many occur closely together, like
$2$ and $3,$ $ 11$ and $13, $ $ 809$ and $811$, on the other hand large gaps
containing no primes also occur. For example, no primes lie between $113$ and
$127, $ $ 293$ and $307, $ $317$ and $331, $ $ 523$ and $541, $ $ 773$ and $787,
$ $ 839$ and $853$ as well as between $887$ and $907$) (For details, see:

{\href{http://www.utm.edu/research/primes/notes/gaps.html}{\tt
http://www.utm.edu/research/primes/notes/gaps.html}}).

This is precisely part of what motivates mathematicians to discover its secrets.
\index{Eratosthenes!sieve}
\paragraph{Sieve of Eratosthenes}
An easy way of calculating all $PI(x)$ primes less than or equal to $x$ is to
use the sieve of Eratosthenes. In the 3rd century before Christ, he found an
extremely easy, automatic way of finding this out. To begin with, you write down
all numbers from 2 to $x$, circle 2, then cross out all multiples of 2. Next,
you circle the lowest number that hasn't been circled or crossed out (3) and
again cross out all multiples of this number, etc. You only need to continue
until you reach the largest number whose square is less than or equal to $x$.

@@

Twin primes are prime numbers whose difference is 2. Examples include 5 and 7 or
101 and 103. Triplet primes, however, only occur once: 3, 5, 7.  For all other
sets of three consecutive uneven numbers, one of them is always divisible by 3
and thus not a prime.
\begin{itemize}
  \item The number of twin primes is an open question: infinite or limited
number?
  The largest twin primes known today are $1,693,965 \cdot 2^{66,443} \pm 1.$
  \item Does a formula exist for calculating the number of twin primes per
interval?
  \item The above proof of the function  $f: N \rightarrow Z$ with $n \mapsto
a^{3^n}$  only guarantees the existence of such a number $a$.  How can we
determine this number $a$ and will it have a value, making the function also of
practical interest?
\item Is there an infinite number of Mersenne prime numbers?
\item Is there an infinite number of Fermat prime numbers?

@@

\paragraph{Further interesting topics regarding prime numbers}
This chapter doesn't consider other number theory topics such as
divisibility rules, modulus calculation, modular inverses, modular powers and
roots, Chinese remainder theorem, Euler PHI function, perfect numbers.

% --------------------------------------------------------------------------
\newpage
\subsubsection{Number of prime numbers in various intervals}\label{s:primhfk}
\hypertarget{primhfk}{}

@@

% --------------------------------------------------------------------------
\newpage
\subsubsection{Indexing prime numbers ($n$-th prime number)} \hypertarget{ntePrimzahl}{}\label{s:ntePrimzahl}

>>>>

\newpage
\section{Prime Numbers}
\hypertarget{Kapitel_2}{}
(Bernhard Esslinger, besslinger@web.de, May 1999, Updates Nov. 2000, Dec. 2001, Feb. 2003)

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid
\footnote{Euclid,
a Greek mathematician of 4th and 3rd century before Christ. He worked at the
Egyptian academy of Alexandria and wrote ``The Elements'', the most well 
known systematically textbook of the Greek mathematics.} 
knew this in the third century before Christ.
\begin{theorem}[Euclid]\label{thm-pz-euklid}\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclidís Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as followed
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than 
any previously existing amount of prime numbers.
} % Ende der Fuﬂnote
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

% --------------------------------------------------------------------------
\hypertarget{MersenneNumbers01}{}
\subsubsection{Special number types -- Mersenne numbers} \index{Mersenne!Mersenne numbers}

@@

\index{Number!Mersenne}\index{Mersenne!Mersenne!number}
\index{Mersenne!theorem}

Even Mersenne knew that not all Mersenne numbers
are prime (see exponent $p = 11$). 
A prime Mersenne number \index{Mersenne!Mersenne prime number} is called Mersenne prime number.  \\
However, he is to be thanked for the interesting conclusion that a number of the form $2^n-1$ is not a prime number
if $n$ is a composite number:

@@

From \begin{eqnarray*} (x^r-1)((x^r)^{s-1} + (x^r)^{s-2} + \cdots + x^r +1) & =
&  ((x^r)^s + (x^r)^{s-1} + (x^r)^{s-2} + \cdots + x^r) \\ &  & -((x^r)^{s-1} +
(x^r)^{s-2} + \cdots + x^r +1)  \\ & = & (x^r)^s -1 = x^{rs } -1,
\end{eqnarray*} we conclude \[ 2^{n_1 n_2} - 1 = (2^{n_1} -1)((2^{n_1})^{n_2 -1}
+ (2^{n_1})^{n_2 -2} + \cdots + 2^{n_1} + 1). \]
Because $ 2^n - 1 $ is a prime number, one of the above two factors on the
right-hand side must be equal to 1. This is the case if and only if $ n_1 =1 $
or $ n_2 =1$. But this contradicts our assumption. Therefore the assumption is
false. This means that there exists no composite number $ n, $ such that $ 2^n -
1 $ is a prime.
\end{Proof}

\vskip + 5pt
\hypertarget{Mer-nums-not-always-prim}{}
Unfortunately this theorem only applies in one direction (the inverse 
statement does not apply, no equivalence): that means that there exist 
prime exponent for which the Mersenne number ist {\bf not} prime (see the 
above example $2^{11}-1, $ where $11$ is prime, but $2^{11}-1$ not).

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before \index{Lucas
  Edouard} Edouard Lucas (1842-1891) proved that this number is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson showed in 1903 which factors make up this composite number: $$ 2^{67} -1
=147, 573, 952, 588, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorisation 
\index{factorisation} (expression as a product of factors) 
\footnote{%
Using CrypTool\index{CrypTool} v1.3 you can factorize numbers in the 
following way: menu {\bf Indiv. Procedures \textbackslash{} RSA Demonstration 
\textbackslash{} Factorisation of a Number}.
}
of this 21-digit decimal number!

Due to the fact that the exponents of the Mersenne numbers
\index{Mersenne!number} do not use all
natural numbers, but only the primes, the {\em experimental space} is limited
considerably. The currently known Mersenne prime numbers 
\index{Mersenne!Mersenne prime number} have the exponents 
$$
\begin{array}{c}
2; ~ 3; ~ 5; ~ 7; ~ 13; ~ 17; ~ 19; ~ 31; ~ 61; ~ 89; ~ 107; ~ 127;
~ 521; ~ 607; ~ 1,279; ~ 2,203; ~ 2,281; ~ 3,217; ~ 4,253;\\
 4,423; ~ 9,689; ~ 9,941, ~ 11,213; ~ 19,937; ~ 21,701; ~ 23,207; ~ 44,497; ~
86,243; ~ 110,503; ~ 132,049; \\
 216,091; ~ 756,839; ~ 859,433; ~ 1,257,787; ~ 1,398,269; ~ 2,976,221; ~ 3,021,377; ~
6,972,593; \\13,466,917.
\end{array}
$$
Thus $39$ Mersenne prime numbers are currently known
\index{Prime number!Mersenne}\index{Mersenne!Mersenne prime number}. 
For the first 38 Mersenne prime numbers we know that this list is complete. The exponents until the 39th Mersenne prime numbers are not yet checked completely.

@@

These numbers can be found at the following URLs:
\begin{itemize}
  \item[] \href{http://reality.sgi.com/chongo/prime/prime_press.html}
               {\texttt{http://reality.sgi.com/chongo/prime/prime\_press.html}} 
  \item[] \href{http://www.utm.edu/}
               {\texttt{http://www.utm.edu/}}
\end{itemize}

\vskip +15 pt
\paragraph{M-37 -- January 1998}\index{Mersenne!Mersenne prime number!M-37}\mbox{}
\vskip +10 pt

The 37th Mersenne prime,
$$ 2^{3,021,377} - 1. $$
was found in January 1998 and has 909,526
digits in the decimal system, which corresponds to 33 pages in the newspaper!

\vskip +15 pt
\paragraph{M-38 -- June 1999}\index{Mersenne!Mersenne prime number!M-38}\mbox{}
\vskip +10 pt

The 38th Mersenne prime, called M-38,
$$ 2^{6,972,593} - 1. $$
was discovered in June 1999 and has $2,098,960$ digits in the decimal system
(that corresponds to around 77 pages in the newspaper).

Right now (Feb. 2003) all exponents smaller than $ 6,972,593 $ have been
checked: so it is sure, that this is really the 38th Mersenne prime number.
This means there is no other Mersenne prime number between the 37th 
Mersenne prime number and this number.

\vskip +15 pt
\paragraph{M13466917 -- December 2001} \index{Mersenne!Mersenne prime number!M-39}\mbox{}
\vskip +10 pt

This number was discovered as 39th Mersenne prime (and already called M-39, despite it is not proven yet, whether there do not exist further Mersenne prime numbers between M-38 und M13466917), $$2^{13,466,917}-1,$$ at December 6, 2001 ñ- more exactly, the verification of this number, found at November 14, 2001 by the Canadian student Michael Cameron, was successfully finished. This number has about 4 million decimal digits (exactly 4,053,946 digits).
Trying only to print this number 
$$(924947738006701322247758 \cdots 1130073855470256259071)$$
would need around 200 pages in the Financial Times.

\vskip +15 pt
\paragraph{GIMPS}\index{GIMPS}\mbox{}
\vskip +10 pt

Discovering the 39th Mersenne prime the GIMPS project (Great Internet Mersenne Prime Search)\index{GIMPS} founded in 1996 already discovered for the 5th time the greatest Mersenne number to be proven being prime
({\href{http://www.mersenne.org} {\tt http://www.mersenne.org}}).

@@

\hypertarget{FermatNumbers01}{}
In the 17th century, Fermat wrote to Mersenne \index{Mersenne Marin} that he presumed that all numbers
of the form $$ F(n) = 2^{2^n} + 1 $$ are prime for all whole numbers $n$ greater
than or equal to $0$ (\hyperlink{FermatNumbers02}{see below}) 
\index{Number!Fermat number}\index{Fermat!Fermat number}

@@

Detailed descriptions of tests for finding out whether a number is prime
can be found on Web sites such as:
\begin{itemize}
  \item[] \href{http://www.utm.edu/research/primes/mersenne.shtml}
               {\texttt{http://www.utm.edu/research/primes/mersenne.shtml}} \\
          \href{http://www.utm.edu/research/primes/prove/index.html}
               {\texttt{http://www.utm.edu/research/primes/prove/index.html}} 
\end{itemize}

@@

Ideas for this can be found at
\begin{itemize}
  \item[] {\href{http://www.utm.edu/research/primes/notes/faq/p_n.html}
          {\tt http://www.utm.edu/research/primes/notes/faq/p\_n.html}}.
\end{itemize}

@@

\item {\em Mersenne numbers}  $f(n) = 2^n - 1$  \quad for $ n $ prime: \\
    As shown \hyperlink{MersenneNumbers01}{above}, this formula seems
    to deliver relatively large prime numbers but - as for $n=11$ 
    [$f(n)=2,047$] - it is repeatedly the case that the result even with
    prime exponents is \hyperlink{Mer-nums-not-always-prim}{not} prime.\\
    Today, all the Mersenne primes having less than 2,000,000 digits are 
    known (M-38) \\
    ({\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}).

@@

    Within the project ``Distributed Search for Fermat Number Dividers''
    offered by Leonid Durman there is also progress in finding new monster
    primes
    ({\href{http://www.fermatsearch.org/}
       {\tt http://www.fermatsearch.org/}}  --  This website has conjectured
    webpages in Russian, Italian and German).

@@

    The Fermat number
    $$ F[2,145,351] = 2^{(2^{2,145,351})} + 1 $$ 
    is divisible by the prime
    $$ p = 3*2^{2,145,353} + 1 $$ \\
    This prime p is the largest known prime generalized Mersenne number. 
    It was discovered only a few days after Michael Angel discovered the 
    largest generalized Fermat number (\hyperlink{generalizedFermatprimes}
    {see below}).  This moved GIMPS' \index{GIMPS} first prime discovery,
    M-35 = M1398269, into 7th place.

@@

    More details are in
    \begin{itemize}
      \item[] \href{http://www.fermatsearch.org/history/cosgrave_record.htm/}
          {\texttt{http://www.fermatsearch.org/history/cosgrave\_record.htm/}}
    \end{itemize}

%    ({\href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
%           {\tt http://perso.wanadoo.fr/yves.gallot/primes/index.html}}).
% M.E. ist die Zahl auf seiner Webseite zu groﬂ, da M-39 "nur" 4 Mio. Stellen hat !
%    Heute kennt man alle Fermatschen Primzahlen mit bis zu 2.000.000.000 ??? 
%    Dezimalstellen. \\

\vskip +15 pt
\hypertarget{generalizedFermatprimes}{}
\item {\em Generalized Fermat numbers}\footnote{%
The base of this power is no more restricted to 2 !}  $F(b,m) = b^{2^m} + 1$:
\index{Fermat!Fermat number!Generalized}  \\
    Generalized Fermat numbers are more numerous than Mersenne numbers at
    equal size and many of them are waiting to be discovered to fill the
    big gaps between the Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the
    representation is not limited to the base 2, can be tested at almost
    the same speed than a Mersenne number.

@@

    Using this program at February 16, 2003 Michael Angel discovered the
    largest of them till then with 628,808 digits, which so became the
    5th largest known prime number:
    $$ b^{2^{17}} + 1  =  62,722^{131,072} + 1. $$ 
    This moved GIMPS' \index{GIMPS} first prime discovery, M-35 = M1398269,
    into 6th place.

    More details are in
    \begin{itemize}
      \item[] \href{http://www.prothsearch.net/index.html}
                   {\texttt{http://www.prothsearch.net/index.html}} \\
              \href{http://perso.wanadoo.fr/yves.gallot/primes/index.html}
               {\texttt{http://perso.wanadoo.fr/yves.gallot/primes/index.html}} 
    \end{itemize}

@@

  \item \index{Euclidean number} {\em Euclidean numbers} $e_n = e_0 
    \cdot e_1 \cdots e_{n-1} + 1$  with $n$ greater than or equal to 
    $1$ and $e_0 := 1$. \\
    $e_{n-1}$ is not the $(n-1)$th prime number, but the number 
    previously found here.
    Unfortunately this formula is not open but recursive.
    The sequence starts with

@@

$e_9, \cdots, e_{17}$ are also composite, which means that this formula is not
particularly useful. Comment: However, what is special about these numbers
is that any pair of them does not have a common factor other than $1$. They are
therefore \index{Prime number!relative}\index{Relatively prim} {\em relatively prime}.

@@

The first $40$ values are prime numbers (which have the obvious regularity that
their difference starts with $2$ and increases by $2$ each time), but the $41$th
and $42$th values are not prime numbers. 
It is easy to see that $f(41)$ cannot be a prime number:
    $f(41) = 41^2 + 41 + 41 = 41 (41 + 1 + 1) = 41 \cdot 43$.

@@

    This means there is no hope to look for further formulas 
    similiar to that of type 11. or 12.

@@

\vskip +10 pt
% --------------------------------------------------------------------------
\subsection{Density and distribution of the primes}

@@

Under the heading {\em How Many Primes Are There} at \\
{\href{http://www.utm.edu/research/primes/howmany.shtml}
{\tt http://www.utm.edu/research/primes/howmany.shtml}} \\
you will find numerous other details.

Using the following Web site: \\
{\href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}
{\tt http://www.math.Princeton.EDU/\~{}arbooker/nthprime.html}} \
you can easily determine $PI(x)$.

The \textbf{distribution} of primes displays several irregularities for which no
ÄsystemÄ has yet been found: On the one hand, many occur closely together, like
$2$ and $3,$ $ 11$ and $13, $ $ 809$ and $811$, on the other hand large gaps
containing no primes also occur. For example, no primes lie between $113$ and
$127, $ $ 293$ and $307, $ $317$ and $331, $ $ 523$ and $541, $ $ 773$ and $787,
$ $ 839$ and $853$ as well as between $887$ and $907$. \\
For details, see: \\
{\href{http://www.utm.edu/research/primes/notes/gaps.html}{\tt
http://www.utm.edu/research/primes/notes/gaps.html}}.

This is precisely part of what motivates mathematicians to discover its secrets.

\index{Eratosthenes!sieve}
\paragraph{Sieve of Eratosthenes}\mbox{}

An easy way of calculating all $PI(x)$ primes less than or equal to $x$ is to
use the sieve of Eratosthenes. In the 3rd century before Christ, he found an
extremely easy, automatic way of finding this out. To begin with, you write down
all numbers from 2 to $x$, circle 2, then cross out all multiples of 2. Next,
you circle the lowest number that hasn't been circled or crossed out (3) and
again cross out all multiples of this number, etc. You only need to continue
until you reach the largest number whose square is less than or equal to $x$.

@@

Twin primes are prime numbers whose difference is 2. Examples include 5 and 7 or
101 and 103. Triplet primes, however, only occur once: 3, 5, 7.  For all other
sets of three consecutive uneven numbers, one of them is always divisible by 3
and thus not a prime.
\begin{itemize}
  \item The number of twin primes is an open question: infinite or limited
number?
  The largest twin primes known today are $1,693,965 \cdot 2^{66,443} \pm 1.$
  \item Does a formula exist for calculating the number of twin primes per
interval?
  \item The above proof of the function  $f: \\qquad N \rightarrow Z$ with $n \mapsto
a^{3^n}$  only guarantees the existence of such a number $a$.  How can we
determine this number $a$ and will it have a value, making the function also of
practical interest?
\item Is there an infinite number of Mersenne prime numbers?\index{Prime number!Mersenne}\index{Mersenne!Mersenne prime number}
\item Is there an infinite number of Fermat prime numbers?

@@

\paragraph{Further interesting topics regarding prime numbers}
\mbox{}

This chapter doesn't consider other number theory topics such as
divisibility rules, modulus calculation, modular inverses, modular powers and
roots, Chinese remainder theorem, Euler PHI function, perfect numbers.
Some of these are considered in the next chapter.

% --------------------------------------------------------------------------
\newpage
\subsubsection{Number of prime numbers in various intervals}\label{s:primhfk}
\hypertarget{primhfk}{}
\vskip +10 pt

@@

% --------------------------------------------------------------------------
\newpage
\subsubsection{Indexing prime numbers ($n$-th prime number)} 
\hypertarget{ntePrimzahl}{}\label{s:ntePrimzahl}
\vskip +10 pt

\begin{tabular}{|l|l|l|l|}\hline
Index   &   Precise value  & Rounded value &
Comment \\
\hline \hline
1       &   2             &     2              & \\
2       &   3             &     3  &  \\
3       &   5             &     5  & \\
4       &   7             &     7 & \\
5       &   11            &     11 & \\
6       &   13            &     13 & \\
7       &   17            &     17 & \\
8       &   19            &     19 & \\
9       &   23            &     23 & \\
10      &   29            &     29 & \\
100     &   541           &     541 & \\
1,000    &   7,917          &     7,917 & \\
664,559  &  9,999,991     &     9.99999E+06 &   All prime numbers up to 1E+07 were known\\
        &                 &                 &  at the beginning of the 20th century. \\
1E+06  &    15,485,863   &      1.54859E+07 & \\
6E+06  &    104,395,301    &    1.04395E+08  & This prime was discovered in 1959.\\
1E+07  &    179,424,673     &    1.79425E+08 & \\
1E+09  &    22,801,763,489  &    2.28018E+10 & \\
1E+12  &    29,996,224,275,833 & 2.99962E+13 & \\ \hline
\end{tabular}
\vskip +2pt Comment: With gaps, extremely large prime numbers were discovered
at an early stage.  \\

\vskip +12pt Web links:

{\href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{http://www.math.Princeton.EDU/\~{}arbooker/nthprime.html}.}

\vskip +16pt Output of the $n$-th prime number

See {\href{http://www.utm.edu/research/primes/notes/by_year.html}
    {\tt http://www.utm.edu/research/primes/notes/by\_year.html}.}
